---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bis620.2023

<!-- badges: start -->
[![R-CMD-check](https://github.com/vavetsbarets/bis620.2023/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/vavetsbarets/bis620.2023/actions/workflows/R-CMD-check.yaml)
[![R-CMD-check](https://github.com/vavetsbarets/bis620.2023/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/vavetsbarets/bis620.2023/actions/workflows/test-coverage.yaml)
<!-- badges: end -->

# Final project BIS620, Vladimir Averin

To access the vignette, click the link (in the about section on the right side) to the webpage. On the webpage open BIS620-final-project article which is the vignette for the final project

## Midterm project Description

bis620.2023 contains the function which plots the spectral signature of the accelerometry data.

At the same time the package contains the shiny app which performs the visual analysis of the clinical trials queried by the user. By running the R/app_project_final.R the user can: 

1. open the app

2. filter the needed trials by the keyword

3. filter the trials by the Sponsor type

4. look at the distribution of the phase of the selected trials

5. observe the dynamics of the number of active trials on the relevant period

6. analyse the conditions/diseases studied in the selected trials

7. look at the adverse events suffered by the subjects of the selected trials

8. observe the geographical distribution of the trials conducted across the world

9. analyse the detailedness of the selected trials

## Installation

You can install the development version of bis620.2023 from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("vavetsbarets/bis620.2023")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(bis620.2023)
accel |> 
  head(100) |> 
  plot_accel()
```

